Fractional Types

### 2019-10-31 fracGC/p.tex
more on intro

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
combination of the Toffoli and Hadamard gates for quantum computing
should not distract from efficiency and safety concerns. The theorems
proving universality (i) assume that temporary storage (often called
\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
this temporary storage is returned to its initial state before
de-allocation. Indeed if no temporary storage is allowed, the Toffoli
gate is not universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as
we demonstrate using space-time tradeoffs in Sec.~\ref{sec:examples},
the more temporary storage is allowed, the more efficient certain
computations could become. The condition requiring that the temporary
storage is only de-allocated when returned to its initial condition is
a safety condition. Violating this condition destroys the symmetry
between input and output making the circuits not reversible and, in
the quantum model, causes irreversible decoherence problems. As
reviewed in Sec.~\ref{sec:examples}, current proposals for reversible
and quantum programming languages
(e.g. Quipper~\cite{Green:2013:QSQ:2491956.2462177}), this safety
condition is at the programmer's discretion whereas.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Motivating Examples}
\label{sec:examples}

@@

In-place matrix transpose: ease of programming, efficiency

Example(s) from Quipper etc. with focus on safety condition

@@

\begin{itemize}

\item Quantum computing hot.

\item Unitary evolution; many algorithms promote classical reversible
functions into unitary gates (examples, including some from Quipper)

\item Common case: given a permutation between finite sets, find a
  reversible program that implements it. Answer depends on the
  available reversible primitives.

\item Example, say we want to transpose a matrix, i.e., write a
  permutation of type \verb|A*B <-> B*A| where \verb|A| and \verb|B|
  are the dimensions. Wikipedia example
  (\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
\begin{verbatim}
11 12 13 14 
21 22 23 24 
\end{verbatim}
  transposed to
\begin{verbatim}
11 21
12 22
13 23
14 24
\end{verbatim}
  Here \verb|A| is \verb|Nat x Nat| and \verb|B| is
  \verb|Nat x Nat x Nat x Nat| and the input and output matrices are:
\begin{verbatim}
M   = (11 , 21) , (12 , 22) , (13, 23) , (14 , 24)

\item So many reversible/quantum circuits allow the introduction of
  so-called ancilla wires, which serve as scratch space (goes back to
  Toffoli). The problem is that there is a discipline for using
  ancilla bits: they need to be created in a known state and can only
  be gc'ed when they are returned to the same state (explain).

\item No language knows how to automatically guarantee this condition;
  left to the programmer (Quipper), or some (incomplete?) dynamic
  checks (Ricercar).

\item We solve this using **fractional types**.

\end{itemize}

Complexity perspective:
\url{https://www.scottaaronson.com/papers/gates.pdf}

Quipper: \url{https://arxiv.org/pdf/1304.3390.pdf}; uses ancilla; one
use is to compile irreversible circuits to reversible ones; need
ancilla bits; more generally several quantum algorithms need ancilla
bits (see below); in quipper de-allocation left to programmer.

Ricercar tries to define rules for ancilla but not complete
\url{https://msoeken.github.io/papers/2015_rc_2.pdf}

>>>>

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
combination of the Toffoli and Hadamard gates for quantum computing
should not distract from efficiency and safety concerns. The theorems
proving universality (i) assume that temporary storage (often called
\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
this temporary storage is returned to its initial state before
de-allocation. Indeed if no temporary storage is allowed, the Toffoli
gate is not universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as
we demonstrate using space-time tradeoffs in Sec.~\ref{sec:examples},
the more temporary storage is allowed, the more efficient certain
computations could become. The condition requiring that the temporary
storage is only de-allocated when returned to its initial condition is
a safety condition. Violating this condition destroys the symmetry
between input and output making the circuits not reversible and, in
the quantum model, causes irreversible decoherence problems. As
reviewed in Sec.~\ref{sec:examples}, ancilla have a number of critical
applications and yet are poorly supported in current reversible and
quantum programming languages making them a common source of bugs.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Ancilla Bits}
\label{sec:examples}

current proposals for reversible
and quantum programming languages
(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
to be used in nested scopes, have sound but incomplete laws for
reasoning about the safety condition, and only guarantee safety via
runtime checks.

@@

In-place matrix transpose: ease of programming, efficiency. Say we
want to transpose a matrix. Wikipedia example
(\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
\begin{verbatim}
11 12 13 14 
21 22 23 24 
\end{verbatim}
transposed to
\begin{verbatim}
11 21
12 22
13 23
14 24
\end{verbatim}
In Pi, the input and output matrices are:
\begin{verbatim}
M   = (11 , 21) , (12 , 22) , (13, 23) , (14 , 24)

Example(s) from Quipper etc. with focus on safety condition. Quipper
uses ancilla bits in several places; one use is to compile
irreversible circuits to reversible ones; need ancilla bits; more
generally several quantum algorithms need ancilla bits (see below); in
quipper de-allocation left to programmer.

@@
